---
title: "CitiBike Desert Analysis"
author: "CJ Robinson"
date: "2025-03-04"
output: html_document
---

## Citi Bike Deserts

This analysis takes data scraped from Citibike stations from August to November 2024 to examine the prevelance of Citi Bike 'deserts' where docks are either full or empty on average of 20% or more of the time. It examines different time periods of morning and evening rushes and midday.

### Set up

```{r setup}
library(tidyverse)
library(lubridate)
library(sf)
library(tigris)
library(gganimate)
library(clipr)


# load in scraped data and mapping data
citibike <- read_csv("../data/citibike_110724.csv")
stations <- read_sf("../data/stations_geo_110724.geojson")


roads_data = primary_secondary_roads("NY") 
landmarks_data = landmarks("NY", type = 'area') 
tracts_data = tracts("NY", county =c("New York", "Kings", "Queens", "Bronx")) 
nj_tracts_data = tracts("NJ", county =c("Bergen", "Hudson")) 

nyc_boroughs <- counties(state = "NY", cb = TRUE, class = "sf") %>%
  filter(NAME %in% c("New York", "Kings", "Queens", "Bronx")) %>% 
  st_transform(., crs = st_crs(stations) )# NYC borough counties 

nj <- counties(state = "NJ", cb = TRUE, class = "sf")  

nyc_water = area_water(state = "NY", county = c("New York", "Kings", "Queens", "Bronx") ) 

nj_water = area_water(state = "NJ", county = c("Bergen", "Hudson") )
```

### Calculates initial analysis

```{r}

hourly_citibike <- citibike %>%
  # get stations in servie
  filter(is_renting == 1, is_returning == 1) %>%
  # cleans data and makes flag variables
  mutate(last_updated = ymd_hms(last_updated),
         in_service = ifelse(num_docks_available == 0 | num_bikes_available + num_ebikes_available == 0, 0, 1),
         docks_available = ifelse(num_docks_available == 0, 0, 1),
         bikes_available = ifelse(num_bikes_available + num_ebikes_available == 0, 0, 1),
         weekend = ifelse(wday(last_updated, week_start = 1) < 6, "weekday", "weekend")) %>%
  # groups by hour/weekend
  group_by(service_hour = hour(last_updated), station_id, weekend) %>%
  # summarize stats
  summarize(
    avg_available = mean(num_docks_available / capacity, na.rm = TRUE),
    returning = mean(docks_available, na.rm = TRUE),
    renting = mean(bikes_available, na.rm = TRUE),
    in_service = mean(in_service, na.rm = TRUE),
    .groups = 'drop'
  ) %>%
  # categorizes based on returning/renting
  mutate(
    rent_return_status = case_when(
      returning <= 0.8 ~ "Full Station (More than 20%)",
      returning <= 0.9 ~ "Full Station (More than 10%)",
      renting <= 0.8 ~ "Empty Station (More than 20%)",
      renting <= 0.9 ~ "Empty Station (More than 10%)",
      .default = "Okay"
    ),
    service_status = case_when(
      in_service > 0.9 ~ "Good (Above 90%)",
      in_service <= 0.9 & in_service >= 0.8 ~ "Okay (Between 90% and 80%)",
      in_service < 0.8 ~ "Bad (Less than 80%)"
    )
  ) %>%  
  left_join(stations, by = "station_id") %>% 
  st_as_sf(sf_column_name = 'geometry') %>%
  st_intersection(nyc_boroughs)  %>%  
  filter(!st_is_empty(geometry))

limits = st_bbox(hourly_citibike$geometry)
```

### Mapping Areas

#### Finding peak times for issues

```{r}
hourly_citibike %>% 
  filter(weekend == "weekday", service_hour >= 5, service_hour <= 20) %>% 
  group_by(service_hour) %>% 
  summarize(returning = 1- mean(returning),
            renting = 1 - mean(renting)) %>% 
  pivot_longer(cols = c("renting", "returning"), names_to = "rent_return", values_to = "percentage") %>% 
  ggplot(aes(x = service_hour, y = percentage)) + 
  geom_bar(stat = "identity") + 
  facet_wrap(~rent_return) +
  scale_x_continuous(breaks = unique(hourly_citibike$service_hour))+
  coord_flip() + 
  theme_minimal()

```

```{r}
hourly_citibike %>% 
  filter(weekend == "weekday", service_hour >= 5, service_hour <= 20) %>% 
  group_by(service_hour) %>% 
  summarize(returning = 1- mean(returning),
            renting = 1 - mean(renting)) %>% 
  pivot_longer(cols = c("renting", "returning"), names_to = "rent_return", values_to = "percentage") %>% 
  ggplot(aes(x = service_hour, y = percentage, fill = rent_return)) + 
  geom_bar(stat = "identity") + 
  theme_minimal()

ggsave("/Documents/github/citi-bike-deserts/img/columns.pdf", width = 375, height = 700, units = 'px')
```


We'll look at 7AM, noon, and 3PM as reference points

#### Entire Map -- 7AM

```{r fig.width=5}
hourly_citibike %>%  
  filter(service_hour == 7, weekend == "weekday") %>%
  ggplot() + 
  geom_sf(data = nyc_boroughs, color = "white", fill = "#FAFAFA") +
  geom_sf(data = tracts_data, color = "white", fill = "#FAFAFA") + 
  geom_sf(data = nyc_water, fill = 'white', color = "white") + 
  geom_sf(data = roads_data, color = '#E6E6E6') + 
  #geom_sf(data = landmarks_data, color = 'white', fill = "#dfede0") + 
  geom_sf(aes(color = service_status), size = .4) +
  scale_color_manual(name = "Average Service Status", values = c("Good (Above 90%)" = "green", "Okay (Between 90% and 80%)" = "orange", "Bad (Less than 80%)" = "red")) +
  coord_sf(xlim = c(as.numeric(limits$xmin),as.numeric(limits$xmax)), 
           ylim = c(as.numeric(limits$ymin),as.numeric(limits$ymax))) +
  theme_void() +
  theme(plot.background = element_rect(fill = 'white', color = 'white')) +
  guides(color = "none")

ggsave("../img/full_city_7am.pdf", width = 5)
```

#### Entire Map -- 12PM

```{r, fig.width = 5}
hourly_citibike %>%  
  filter(service_hour == 12, weekend == "weekday") %>%
  ggplot() + 
  geom_sf(data = nyc_boroughs, color = "white", fill = "#FAFAFA") +
  geom_sf(data = tracts_data, color = "white", fill = "#FAFAFA") + 
  geom_sf(data = nyc_water, fill = 'white', color = "white") + 
  geom_sf(data = roads_data, color = '#E6E6E6') + 
  #geom_sf(data = landmarks_data, color = 'white', fill = "#dfede0") + 
  geom_sf(aes(color = service_status), size = .4) +
  scale_color_manual(name = "Average Service Status", values = c("Good (Above 90%)" = "green", "Okay (Between 90% and 80%)" = "orange", "Bad (Less than 80%)" = "red")) +
  coord_sf(xlim = c(as.numeric(limits$xmin),as.numeric(limits$xmax)), 
           ylim = c(as.numeric(limits$ymin),as.numeric(limits$ymax))) +
  theme_void() +
  theme(plot.background = element_rect(fill = 'white', color = 'white')) +
  guides(color = "none")



ggsave("../img/full_city_12pm.pdf", width = 5)
```

#### Entire Map -- 3PM

```{r, fig.width = 5}
hourly_citibike %>%  
  filter(service_hour == 15, weekend == "weekday") %>%
  ggplot() + 
  geom_sf(data = nyc_boroughs, color = "white", fill = "#FAFAFA") +
  geom_sf(data = tracts_data, color = "white", fill = "#FAFAFA") + 
  geom_sf(data = nyc_water, fill = 'white', color = "white") + 
  geom_sf(data = roads_data, color = '#E6E6E6') + 
  #geom_sf(data = landmarks_data, color = 'white', fill = "#dfede0") + 
  geom_sf(aes(color = service_status), size = .4) +
  scale_color_manual(name = "Average Service Status", values = c("Good (Above 90%)" = "green", "Okay (Between 90% and 80%)" = "orange", "Bad (Less than 80%)" = "red")) +
  coord_sf(xlim = c(as.numeric(limits$xmin),as.numeric(limits$xmax)), 
           ylim = c(as.numeric(limits$ymin),as.numeric(limits$ymax))) +
  theme_void() +
  theme(plot.background = element_rect(fill = 'white', color = 'white')) +
  guides(color = "none")

ggsave("../img/full_city_3pm.pdf", width = 5)

```

----

### East Village 

```{r}
hourly_citibike %>%  
  filter(service_hour == 7, weekend == "weekday") %>%
ggplot() + 
  geom_sf(data = nyc_boroughs, color = "white", fill = "#FAFAFA") +
  geom_sf(data = tracts_data, color = "white", fill = "#FAFAFA") + 
  geom_sf(data = nyc_water, fill = 'lightblue', color = "white") + 
  geom_sf(data = roads_data, color = '#E6E6E6') + 
  geom_sf(aes(color = service_status), size = 1) +
  scale_color_manual(name = "Average Service Status", 
                     values = c("Good (Above 90%)" = "green", 
                                "Okay (Between 90% and 80%)" = "orange", 
                                "Bad (Less than 80%)" = "red")) +
coord_sf(xlim = c(-73.99, -73.97), 
        ylim = c(40.72, 40.74), 
        expand = FALSE) +
  theme_void() +
  theme(
    plot.background = element_rect(fill = 'white', color = 'white'),
    aspect.ratio = 1
  ) +
  guides(color = "none")

ggsave("../img/evillage_7am.png", width = 200, height = 200, units = 'px')


hourly_citibike %>%  
  filter(service_hour == 12, weekend == "weekday") %>%
ggplot() + 
  geom_sf(data = nyc_boroughs, color = "white", fill = "#FAFAFA") +
  geom_sf(data = tracts_data, color = "white", fill = "#FAFAFA") + 
  geom_sf(data = nyc_water, fill = 'lightblue', color = "white") + 
  geom_sf(data = roads_data, color = '#E6E6E6') + 
  geom_sf(aes(color = service_status), size = 1) +
  scale_color_manual(name = "Average Service Status", 
                     values = c("Good (Above 90%)" = "green", 
                                "Okay (Between 90% and 80%)" = "orange", 
                                "Bad (Less than 80%)" = "red")) +
coord_sf(xlim = c(-73.99, -73.97), 
        ylim = c(40.72, 40.74), 
        expand = FALSE) +
  theme_void() +
  theme(
    plot.background = element_rect(fill = 'white', color = 'white'),
    aspect.ratio = 1
  ) +
  guides(color = "none")

ggsave("../img/evillage_12pm.png", width = 200, height = 200, units = 'px')


hourly_citibike %>%  
  filter(service_hour == 15, weekend == "weekday") %>%
ggplot() + 
  geom_sf(data = nyc_boroughs, color = "white", fill = "#FAFAFA") +
  geom_sf(data = tracts_data, color = "white", fill = "#FAFAFA") + 
  geom_sf(data = nyc_water, fill = 'lightblue', color = "white") + 
  geom_sf(data = roads_data, color = '#E6E6E6') + 
  geom_sf(aes(color = service_status), size = 1) +
  scale_color_manual(name = "Average Service Status", 
                     values = c("Good (Above 90%)" = "green", 
                                "Okay (Between 90% and 80%)" = "orange", 
                                "Bad (Less than 80%)" = "red")) +
coord_sf(xlim = c(-73.99, -73.97), 
        ylim = c(40.72, 40.74), 
        expand = FALSE) +
  theme_void() +
  theme(
    plot.background = element_rect(fill = 'white', color = 'white'),
    aspect.ratio = 1
  ) +
  guides(color = "none")

ggsave("../img/evillage_3pm.png", width = 200, height = 200, units = 'px')


```



----

### Midtown

```{r}
hourly_citibike %>%  
  filter(service_hour == 7, weekend == "weekday") %>%
ggplot() + 
  geom_sf(data = nyc_boroughs, color = "white", fill = "#FAFAFA") +
  geom_sf(data = tracts_data, color = "white", fill = "#FAFAFA") + 
  geom_sf(data = nyc_water, fill = 'white', color = "white") + 
  geom_sf(data = roads_data, color = '#E6E6E6') + 
  geom_sf(data = landmarks_data, color = 'white', fill = "#dfede0") + 

  geom_sf(aes(color = service_status), size = 1) +
  scale_color_manual(name = "Average Service Status", 
                     values = c("Good (Above 90%)" = "green", 
                                "Okay (Between 90% and 80%)" = "orange", 
                                "Bad (Less than 80%)" = "red")) +
  coord_sf(xlim = c(-73.985, -73.965), 
         ylim = c(40.75, 40.77), 
         expand = FALSE) +
  theme_void() +
  theme(
    plot.background = element_rect(fill = 'white', color = 'white'),
    aspect.ratio = 1
  ) +
  guides(color = "none")

ggsave("../img/midtown_7am.png", width = 200, height = 200, units = 'px')


hourly_citibike %>%  
  filter(service_hour == 12, weekend == "weekday") %>%
ggplot() + 
  geom_sf(data = nyc_boroughs, color = "white", fill = "#FAFAFA") +
  geom_sf(data = tracts_data, color = "white", fill = "#FAFAFA") + 
  geom_sf(data = nyc_water, fill = 'white', color = "white") + 
  geom_sf(data = roads_data, color = '#E6E6E6') + 
  geom_sf(data = landmarks_data, color = 'white', fill = "#dfede0") + 
  geom_sf(aes(color = service_status), size = 1) +
  scale_color_manual(name = "Average Service Status", 
                     values = c("Good (Above 90%)" = "green", 
                                "Okay (Between 90% and 80%)" = "orange", 
                                "Bad (Less than 80%)" = "red")) +
  coord_sf(xlim = c(-73.985, -73.965), 
         ylim = c(40.75, 40.77), 
         expand = FALSE) +
  theme_void() +
  theme(
    plot.background = element_rect(fill = 'white', color = 'white'),
    aspect.ratio = 1
  ) +
  guides(color = "none")

ggsave("../img/midtown_12pm.png", width = 200, height = 200, units = 'px')


hourly_citibike %>%  
  filter(service_hour == 15, weekend == "weekday") %>%
ggplot() + 
  geom_sf(data = nyc_boroughs, color = "white", fill = "#FAFAFA") +
  geom_sf(data = tracts_data, color = "white", fill = "#FAFAFA") + 
  geom_sf(data = nyc_water, fill = 'white', color = "white") + 
  geom_sf(data = landmarks_data, color = 'white', fill = "#dfede0") + 
  geom_sf(data = roads_data, color = '#E6E6E6') + 
  geom_sf(aes(color = service_status), size = 1) +
  coord_sf(xlim = c(-73.985, -73.965), 
         ylim = c(40.75, 40.77), 
         expand = FALSE) +
  scale_color_manual(name = "Average Service Status", 
                     values = c("Good (Above 90%)" = "green", 
                                "Okay (Between 90% and 80%)" = "orange", 
                                "Bad (Less than 80%)" = "red")) +
  theme_void() +
  theme(
    plot.background = element_rect(fill = 'white', color = 'white'),
    aspect.ratio = 1
  ) +
  guides(color = "none")

ggsave("../img/midtown_3pm.pdf", width = 200, height = 200, units = 'px')
ggsave("../img/midtown_3pm.png", width = 200, height = 200, units = 'px')


```

------

#### W Village

```{r}
hourly_citibike %>%  
  filter(service_hour == 7, weekend == "weekday") %>%
ggplot() + 
  geom_sf(data = nyc_boroughs, color = "white", fill = "#FAFAFA") +
  geom_sf(data = tracts_data, color = "white", fill = "#FAFAFA") + 
  geom_sf(data = nyc_water, fill = 'lightblue', color = "white") + 
  geom_sf(data = roads_data, color = '#E6E6E6') + 
  geom_sf(data = landmarks_data, color = 'white', fill = "#dfede0") + 

  geom_sf(aes(color = service_status), size = 1) +
  scale_color_manual(name = "Average Service Status", 
                     values = c("Good (Above 90%)" = "green", 
                                "Okay (Between 90% and 80%)" = "orange", 
                                "Bad (Less than 80%)" = "red")) +
coord_sf(xlim = c(-74.02, -74.00), 
         ylim = c(40.71, 40.73), 
         expand = FALSE) +
  theme_void() +
  theme(
    plot.background = element_rect(fill = 'white', color = 'white'),
    aspect.ratio = 1
  ) +
  guides(color = "none")

ggsave("../img/wvillage_7am.png", width = 200, height = 200, units = 'px')


hourly_citibike %>%  
  filter(service_hour == 12, weekend == "weekday") %>%
ggplot() + 
  geom_sf(data = nyc_boroughs, color = "white", fill = "#FAFAFA") +
  geom_sf(data = tracts_data, color = "white", fill = "#FAFAFA") + 
  geom_sf(data = nyc_water, fill = 'lightblue', color = "white") + 
  geom_sf(data = roads_data, color = '#E6E6E6') + 
  geom_sf(data = landmarks_data, color = 'white', fill = "#dfede0") + 
  geom_sf(aes(color = service_status), size = 1) +
  scale_color_manual(name = "Average Service Status", 
                     values = c("Good (Above 90%)" = "green", 
                                "Okay (Between 90% and 80%)" = "orange", 
                                "Bad (Less than 80%)" = "red")) +
coord_sf(xlim = c(-74.02, -74.00), 
         ylim = c(40.71, 40.73), 
         expand = FALSE) +
  theme_void() +
  theme(
    plot.background = element_rect(fill = 'white', color = 'white'),
    aspect.ratio = 1
  ) +
  guides(color = "none")

ggsave("../img/wvillage_12pm.png", width = 200, height = 200, units = 'px')


hourly_citibike %>%  
  filter(service_hour == 15, weekend == "weekday") %>%
ggplot() + 
  geom_sf(data = nyc_boroughs, color = "white", fill = "#FAFAFA") +
  geom_sf(data = tracts_data, color = "white", fill = "#FAFAFA") + 
  geom_sf(data = nyc_water, fill = 'lightblue', color = "white") + 
  geom_sf(data = landmarks_data, color = 'white', fill = "#dfede0") + 
  geom_sf(data = roads_data, color = '#E6E6E6') + 
  geom_sf(aes(color = service_status), size = 1) +
coord_sf(xlim = c(-74.02, -74.00), 
         ylim = c(40.71, 40.73), 
         expand = FALSE) +
  scale_color_manual(name = "Average Service Status", 
                     values = c("Good (Above 90%)" = "green", 
                                "Okay (Between 90% and 80%)" = "orange", 
                                "Bad (Less than 80%)" = "red")) +
  theme_void() +
  theme(
    plot.background = element_rect(fill = 'white', color = 'white'),
    aspect.ratio = 1
  ) +
  guides(color = "none")

ggsave("../img/wvillage_3pm.png", width = 200, height = 200, units = 'px')


```

